---
title: "Drivers of Urban Housing Prices"
subtitle: "Analyzing Population, Income, and Housing Demand Dynamics"
author: 
  - Rui Hu
thanks: "Code and data are available at: [https://github.com/XYPKQ896/housing_price](https://github.com/XYPKQ896/housing_price)."
date: 27 November 2024
date-format: long
abstract: "This paper examines the factors influencing urban housing prices, focusing on demographic, economic, and market dynamics. We analyze key drivers such as population growth, household characteristics, income levels, housing demand, and rental prices. Our findings reveal how these variables interact to shape housing market trends and affordability. Understanding these relationships provides valuable insights for policymakers and stakeholders to address urban housing challenges and improve housing accessibility."
format: pdf
number-sections: true
toc: true
toc-title: "Table of Contents" 
toc-depth: 3 
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(arrow)
library(broom)
library(knitr)
data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
```


# Introduction

Urban housing prices are shaped by a complex interplay of demographic, economic, and market factors. Cities worldwide face unique challenges as population growth, changing household structures, income levels, and rental market dynamics influence housing demand and affordability. Understanding these variables is essential for unraveling the underlying mechanisms of urban housing price fluctuations and addressing persistent housing market issues.

This study aims to estimate the relative influence of six critical factors—population size, household size, number of households, monthly income, housing upgrade demand, and rental prices—on urban housing prices. By analyzing these variables, we identify their individual contributions and interactions in shaping housing markets.

The analysis reveals that several factors significantly influence the average market price of housing. Population size has a negative impact on housing prices, with a substantial decrease in price associated with population growth. Additionally, both the average household size and the number of households are positively related to housing prices, with larger household sizes and more households driving up market prices. Monthly income also has a positive effect on housing prices, indicating that higher income levels are associated with higher housing prices.

These insights are crucial for policymakers and market stakeholders to design targeted interventions that promote housing affordability, stability, and equity in urban areas. By addressing the root causes of housing market pressures, this research contributes to the broader goal of creating sustainable and inclusive cities.

Telegraphing paragraph: The remainder of this paper is structured as follows. 
@sec-data Data, @sec-model Model, @sec-results Results, @sec-discussion Discussion.


# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR] for data analysis. 
We use ggplot2 @ggplot2 to visualize the data. We use arrow @R-arrow to read the parquet.
The data used in this study comes from the website AIDDATA [@aiddata]. The article 
[@costs_City] includes national census data, housing market surveys, and rental market reports. Following the guidelines in Telling Stories with Data 
[@tellingstories], we focus on key variables such as population size, household size, number of households, monthly income, housing upgrade demand, and rental prices to capture the multifaceted nature of urban housing markets. The dataset spans urban areas in various regions, providing a broad perspective on housing price dynamics.

The preview of the original and cleaned dataset can be found in Appendix. @sec-additional_data

## Measurement
	
To translate real-world phenomena into dataset entries, we employ standardized metrics and definitions. Population size is measured as the total number of residents in an urban area, while household size represents the average number of people per household, derived from census microdata. The number of households is computed by dividing the total population by the household size.
Monthly income is represented by low quantile income at the city level, adjusted for inflation and regional purchasing power. Housing upgrade demand is quantified as the percentage of households expressing a preference for moving to larger or higher-quality homes, based on survey responses. Finally, rental prices reflect the average monthly rent of an adequate home 550 SqFt per month, serving as a proxy for rental market trends.
Data cleaning and validation processes ensure consistency across variables, and missing values are imputed using multiple imputation techniques to minimize bias.

## Outcome variables


### Market Prices
Market prices @fig-market_prices are represented by the average market price of an adequate home of 550 sqft in each urban area. This variable is a direct measure of urban housing costs and reflects the interaction of supply and demand in the housing market.

```{r, fig.pos='H'}
#| label: fig-market_prices
#| fig-cap: Market Prices
#| echo: false
#| warning: false
#| message: false
ggplot(data, aes(x = Average_Market_Price_of_an_Adequate_Home)) +
  geom_histogram(binwidth = 40000) +
  labs(
    title = "Figure 1: Distribution of Market Prices",
    x = "Market Price (in USD)",
    y = "Frequency"
  ) +
  theme_minimal()
```
Most of market price of our data is smaller than 10000 USD. It shows a right 
skewed which displays that a few luxury homes with significantly higher prices.

## Predictor variables


### Population
Some of our data is of population (@fig-population).
The total population of each city, which influences demand for housing and public services. Larger populations are often correlated with higher demand for housing and more complex governance challenges.

```{r, fig.pos='H'}
#| label: fig-population
#| fig-cap: Population across cities.
#| echo: false
ggplot(data = data, aes(x = City, y = Population)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Figure 2: Population Size Across Cities", x = "City", y = "Population Size") +
  theme(axis.text.x = element_blank()) 
```

This variable shows the distribution of population across different cities, revealing notable variation between them. The city with the largest population is significantly larger than the others.


### Size
The overall size of a city (@fig-size) is a measure of the geographical area or population density of an urban area. It is an important factor influencing housing markets, as larger cities often have higher demand for housing due to more concentrated economic opportunities and services. City size was derived from official urban area boundaries and population density data, typically measured in square miles or kilometers, and is often correlated with higher housing prices due to demand-supply imbalances.
```{r, fig.pos='H'}
#| label: fig-size
#| fig-cap: Size across cities.
#| echo: false
ggplot(data, aes(x = Size)) +
  geom_bar() +
  labs(title = "Figure 3: Distribution of City Size Categories",
       x = "City Size Category", 
       y = "Frequency") +
  theme_minimal()

```
Medium-sized cities have the highest frequency, appearing significantly more frequently than the "Large" and "Small" categories.
Large cities appear the least frequently, with fewer cities falling into this category compared to the other two.
Small cities show a moderate frequency, falling between the other two categories.


### Average Household Size
Average Household Size represents the average number of individuals living in a single household. Smaller household sizes often lead to an increased number of households, elevating housing demand. Calculated directly from census microdata, this measure captures changes in household formation trends.
```{r, fig.pos='H'}
#| echo: false
# Calculate summary statistics for Average_HH_Size
summary_stats <- summary(data$Average_HH_Size)

# Convert the summary output into a data frame for kable
summary_stats_df <- as.data.frame(matrix(summary_stats, ncol = 1, dimnames = list(names(summary_stats), "Value")))

# Render the summary statistics table
summary_stats_df %>% kable(caption = "Summary Statistics for Average Household Size")
```
In conclusion, these statistics suggest that urban areas in the dataset tend to have moderately sized households, with a slight tendency towards larger household sizes in the upper quartile.


### Number of Households
Number of Households @fig-number_of_HH is derived by dividing the total population by the average household size. It serves as a direct measure of housing unit demand in each urban area.
```{r, fig.pos='H'}
#| label: fig-number_of_HH
#| fig-cap: Number of Households.
#| echo: false
#| message: false
#| warning: false
ggplot(data, aes(x = Population, y = Number_of_HH)) +
  geom_point() +  
  geom_smooth(method = "lm") + 
  labs(
    title = "Figure 4: Relationship between Population Size and Number of Households",
    x = "Population",
    y = "Number of Households"
  ) +
  theme_minimal() 

```
This plot confirms that population size is a strong predictor of the number of households in urban areas. It validates the expectation that larger populations generally lead to more households, which directly impacts housing demand and urban planning. However, the variability in the relationship at higher population sizes suggests that additional factors, such as average household size or regional housing policies, might play a role and could be investigated further.


### Monthly Income
Monthly income @fig-Monthly_Income is the average monthly household income of the lowest quintile of Population in each urban area. This variable is critical for assessing housing affordability and purchasing power in the market.

```{r, fig.pos='H'}
#| label: fig-Monthly_Income
#| fig-cap: Monthly Income.
#| echo: false
#| message: false
#| warning: false
ggplot(data, aes(x = Monthly_Income)) +
  geom_histogram(binwidth = 8000) +
  labs(
    title = "Figure 5: Distribution of Monthly Income Across Cities",
    x = "Monthly Income (in USD)",
    y = "Frequency"
  ) +
  theme_minimal() 

```
The histogram shows that most monthly incomes are clustered on the lower end (below $5,000).
The distribution is right-skewed, indicating a smaller number of individuals or households with significantly higher incomes.
Monthly incomes range widely, from near $0 to over $15,000. This wide range signifies diverse economic conditions across cities, potentially driven by city size, industrial composition, or cost of living.


### Percentage of Households Needing Housing Upgrades
Percentage of Households Needing Housing Upgrades @fig-upgrades reflects the proportion of households expressing a desire to move to larger or higher-quality homes. It captures the aspirational demand for improved housing conditions, which can drive housing market activity. Survey data were used to compute these percentages at the city level.
```{r, fig.pos='H'}
#| label: fig-upgrades
#| fig-cap: Monthly Income.
#| echo: false
#| message: false
#| warning: false
summary_table <- data %>%
  group_by(Size) %>%
  summarise(
    Total_Households = sum(Number_of_HH, na.rm = TRUE),
    percentage_Needing_Upgrades = mean(Percentage_of_HH_in_Need_of_Adequate_Housing_Upgrades)
  ) %>%
  mutate(Needing_Upgrades = Total_Households * percentage_Needing_Upgrades)

# Create a formatted table
knitr::kable(
  summary_table,
  col.names = c("Size", "Total Households", "Percentage Needing Upgrades (%)", "Households Needing Upgrades"),
  caption = "Percentage of Households Needing Housing Upgrades by Size",
  digits = 2
)
```
Households in the small cities are willing to upgrades their home. Households in the medium city has a smaller
preference to upgrade the home.


### Montly Rent
Monthly rent @fig-Monthly_rent means the average monthly rent of an adequate home 550 SqFt per month. This measure captures housing demand and serves as a proxy for housing affordability.
```{r, fig.pos='H'}
#| label: fig-Monthly_rent
#| fig-cap: Monthly Income.
#| echo: false
#| message: false
#| warning: false
ggplot(data, aes(x = Average_Monthly_Rent_of_an_Adequate_Home)) +
  geom_histogram(binwidth = 2000) +
  labs(
    title = "Figure 7: Distribution of Monthly Rent Across Cities",
    x = "Monthly Rent (in USD)",
    y = "Frequency"
  ) +
  theme_minimal() 

```
The distribution is heavily right-skewed, with a majority of values concentrated at the lower end of the range.
A large number of cities have monthly rents clustered around $0–$1250, indicating relatively affordable housing in many areas.


# Model{#sec-model}

The goal of our modeling strategy is twofold. Firstly, we aim to quantify the relationships between various demographic and housing market characteristics and the average market price of an adequate home. Secondly, we use this model to identify key predictors that could inform housing policy and urban planning. The chosen model combines demographic, economic, and housing variables to provide a comprehensive understanding of market dynamics. Background details and diagnostics are included in [Appendix - @sec-model-details].

## Model set-up

Define $y_i$ as the average market price of an adequate home for city $i$, measured in USD. The predictor variables include:

- $x_{1i}$: Population size of city $i$.
- $x_{2i}$: Size of city $i$ (e.g., small, medium, or large).
- $x_{3i}$: Average household size in city $i$.
- $x_{4i}$: Total number of households in city $i$.
- $x_{5i}$: Average monthly income in USD for city $i$.
- $x_{6i}$: Percentage of households needing adequate housing upgrades in city $i$.
- $x_{7i}$: Average monthly rent for an adequate home in city $i$.

The linear regression model is defined as:

$y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \beta_3 x_{3i} + \beta_4 x_{4i} + \beta_5 x_{5i} + \beta_6 x_{6i} + \beta_7 x_{7i}$

where:
- $\beta_0$ is the intercept.

- $\beta_1, \dots, \beta_7$ are the coefficients corresponding to the predictors.

We run the model in R [@citeR] using the `lm()` function from the base R package.


### Model justification

The predictors were selected based on their theoretical relevance to housing market prices. Key assumptions include:

1. Population ($x_{1i}$): Larger populations may increase housing demand, driving prices higher.
2. City size ($x_{2i}$): Larger cities often experience higher housing prices due to limited land and increased demand.
3. Household size ($x_{3i}$): Larger households may influence the demand for larger homes, potentially increasing prices.
4. Number of households ($x_{4i}$): This variable captures the housing supply-demand balance, which impacts pricing.
5. Income ($x_{5i}$): Cities with higher average incomes often experience greater purchasing power, increasing housing demand.
6. Housing upgrade needs ($x_{6i}$): A higher percentage of inadequate housing could depress prices if supply quality is low.
7. Rent ($x_{7i}$): Rent prices often correlate with home prices, acting as a proxy for housing affordability.


# Results{#sec-results}

Our results are summarized in @tbl-modelresults which presents the coefficients of the explanatory model predicting the average market price of an adequate home. @broom is used to tidy the model. @knitr is used to kable the table.
The statistically significant is p<0.05.
The model reveals several important insights and indicates a baseline average market price of $40,500 when all predictor variables are at their reference levels or zero. Among the predictor variables:


```{r, fig.pos="h"}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of average market price"
#| warning: false

# Create a summary table of the model results
model1 <- readRDS(here::here("models/second_model.rds"))

model_results <- tidy(model1) %>%
  mutate(
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "Population" ~ "Population",
      term == "SizeMedium" ~ "Size: Medium (Ref: Large)",
      term == "SizeSmall" ~ "Size: Small (Ref: Large)",
      term == "Average_HH_Size" ~ "Average Household Size",
      term == "Number_of_HH" ~ "Number of Households",
      term == "Monthly_Income" ~ "Monthly Income",
      term == "Percentage_of_HH_in_Need_of_Adequate_Housing_Upgrades" ~ "Percentage of Households Needing Upgrades",
      term == "Average_Monthly_Rent_of_an_Adequate_Home" ~ "Average Monthly Rent",
      TRUE ~ term
    )
  )

# Format the table
kable(
  model_results[, c("term", "estimate", "std.error", "statistic", "p.value")],
  col.names = c("Predictor", "Estimate", "Std. Error", "t value", "Pr(>|t|)"),
  digits = 3
)

```
Intercept: The intercept has an estimate of -1592.501 (p=0.002), suggesting a baseline adjustment for average market prices when all predictors are held constant.

Population: The coefficient for population is -11123.163 (p=0.002), indicating that increases in population are associated with a significant negative impact on market prices, possibly reflecting supply pressures or urban constraints.

City Size:
Medium cities (compared to large cities) have an insignificant positive estimate of 0.057 (p=0.230).
Small cities have a negligible and statistically insignificant estimate of -0.001 (p=0.986), indicating that city size may not strongly predict market prices when controlling for other factors.

Average Household Size: With an estimate of 764.739 (p=0.002), this variable significantly impacts market prices positively, suggesting that larger household sizes may be associated with higher market values.

Number of Households: The coefficient is 11946.141 (p=0.002), showing a significant positive association between the number of households and market prices, possibly reflecting demand pressures.

Monthly Income: With an estimate of 0.696 (p=0.003), higher monthly income is significantly associated with higher market prices, consistent with economic theory linking income levels to housing affordability.

Percentage of Households Needing Upgrades: This variable has an estimate of 0.085 (p=0.182), suggesting a positive but statistically insignificant relationship.

Average Monthly Rent: The coefficient is 0.547 (p=0.078), which approaches significance, indicating that higher rents may predict higher market prices.




# Discussion{#sec-discussion}

## The Relationship Among Population and Market Price {#sec-first-point}

The analysis indicates a significant negative relationship between population size and average market prices for adequate homes ($\beta$=−11,123.163,p=0.002). This finding suggests that, contrary to conventional wisdom, larger populations are associated with lower average housing prices.

Population size is often linked to urban agglomeration, where economic opportunities attract a broad spectrum of residents. While demand increases, cities with larger populations may also benefit from broader tax bases and investment in affordable housing initiatives that temper price growth.

## The Relationship Between Average Household Size and Market Price

The analysis reveals a statistically significant positive relationship between average household size and the average market price for adequate homes ($\beta$=764.739,p=0.002). This finding suggests that, as the average household size increases, so does the market price for housing in that area.

Larger households often require more space, which can drive up demand for bigger homes. In many urban areas, larger households may have a higher propensity to purchase or rent larger, more expensive homes, thereby increasing the average market price.

## The Relationship Between Number of Household and Market Price

The analysis shows a statistically significant positive relationship between the number of households and the average market price for adequate homes ($\beta$=11,946.141,p=0.002). This suggests that as the number of households increases, the average market price of homes also tends to rise.

As the number of households in an area grows, the demand for housing naturally increases. More households mean a greater need for available housing units, which can push prices up, especially in areas where supply is limited.
A larger number of households creates greater competition for the available housing stock. This heightened competition can lead to higher prices, particularly in areas with constrained housing supply. When the number of households exceeds the number of available units, bidding wars or higher rents can drive up market prices.

## The Relationship Between Monthly Income and Market Price
The analysis reveals a statistically significant positive relationship between monthly income and the average market price of homes ($\beta$=0.696,p=0.003). This indicates that as monthly income increases, the market price for adequate homes tends to rise.

Higher monthly income increases individuals' purchasing power, allowing them to afford more expensive homes. As income levels rise, people are more willing and able to bid for higher-quality homes, which in turn pushes up the average market price.
In regions with higher average monthly incomes, there is often a demand for larger, more luxurious homes. This is particularly true in affluent urban centers where wealthier individuals tend to live, which raises the overall market price. Such areas are more likely to experience an increase in the availability of higher-end properties, contributing to a higher average home price.

## Weaknesses and next steps

Despite the valuable insights provided by the model, there are several weaknesses that limit the comprehensiveness and robustness of the analysis. Additionally, several next steps can be taken to address these limitations and improve our understanding of housing market dynamics.

Weaknesses
Model Simplification and Omitted Variables
One significant weakness of the current model is its simplification. While it includes several key predictors such as Monthly Income, Monthly Rent, and Household Size, there may be other important variables influencing housing prices that were not included in the model. Factors such as interest rates, government policies, zoning laws, and infrastructure development are likely to play a significant role in shaping housing prices. The omission of these variables could lead to omitted variable bias, potentially distorting the model’s findings. The lack of certain unobserved variables, such as neighborhood-level amenities, crime rates, and future growth projections, also limits the model’s predictive power.

Next Steps
Incorporating Additional Variables
To address the limitation of omitted variables, future research should aim to incorporate additional predictors such as interest rates, tax policies, neighborhood-level amenities, and zoning regulations. These factors are often highly correlated with housing prices and would provide a more comprehensive understanding of what drives housing markets. Moreover, including time-varying factors could capture the evolving nature of housing markets.

In conclusion, while this model provides valuable insights into the relationship between income, rent, and housing prices, there are several limitations that should be addressed in future research. By incorporating more data, addressing multicollinearity, and improving the model’s assumptions, we can create a more robust and comprehensive framework for understanding housing market dynamics.

\newpage

\appendix

# Appendix {-}


# Additional data details {#sec-additional_data}
Preview of the raw dataset
```{r, fig.pos="h"}
#| label: tbl-raw
#| tbl-cap: "Preview of the raw dataset"
#| message: false
#| echo: false
raw <- read_csv(here::here("data/01-raw_data/original_data.csv"))
head(raw)[1:5] %>% kable()
head(raw)[6:7] %>% kable()
head(raw)[8:9] %>% kable()
head(raw)[10:13] %>% kable()
head(raw)[14:16] %>% kable()
head(raw)[17:20] %>% kable()
head(raw)[21:23] %>% kable()
head(raw)[24:25] %>% kable()
head(raw)[26:27] %>% kable()
head(raw)[28:30] %>% kable()
head(raw)[31:31] %>% kable()
head(raw)[32:34] %>% kable()
head(raw)[34:36] %>% kable()
head(raw)[37:38] %>% kable()
```
Preview of the cleaned dataset
```{r, fig.pos="h"}
#| label: tbl-cleaned
#| tbl-cap: "Preview of the raw dataset"
#| message: false
#| echo: false
head(data)[1:5] %>% kable()
head(data)[6:7] %>% kable()
head(data)[8:8] %>% kable()
head(data)[9:9] %>% kable()
head(data)[10:10] %>% kable()
```

# Model details {#sec-model-details}

## Check the violation assumptions of original model.

In @fig-first-1 we implement a residual vs. fitted value plot. This shows clustering of residuals in certain ranges of fitted values, with uneven dispersion in others. This violates constant variance assumption.

In @fig-first-2 we draw a QQ plot. This shows the deviation at the extremes indicates the presence of outliers or non-normality in the residuals, which could bias parameter estimates and predictions.It violates normality assumption.

```{r, fig.pos='H'}
#| echo: false
#| message: false
#| warning: false
#| label: fig-first
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-pos: "h"
model <- readRDS(here::here("models/first_model.rds"))
plot(model$fitted.values, resid(model), 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     main = "Residuals vs Fitted")
qqnorm(resid(model))
qqline(resid(model))
```

## Check the violation assumptions of transfered model.
By applying Box-cox transformation to the first model. We get the second model.
Now, we want to check the assumptions of second model.

@fig-second-1 is a Residual vs. fited value plot. It shows that The residuals appear to be randomly distributed around zero, without a clear pattern. Also, the spread of residuals does not seem to systematically increase or decrease with the fitted values.
This suggests that the model does not violate uncorrelated, linearity or  constant variance assumptions.

@fig-second-2 is a QQ plot. It shows that all the points are located near the line.

This suggests that the model does not violate normality assumption.

```{r, fig.pos='H'}
#| echo: false
#| message: false
#| warning: false
#| label: fig-second
#| fig-cap: "Checking the second model"
#| fig-pos: "h"
model1 <- readRDS(here::here("models/second_model.rds"))
plot(model1$fitted.values, resid(model1), 
     xlab = "Fitted Values", 
     ylab = "Residuals", 
     main = "Residuals vs Fitted")
qqnorm(resid(model1))
qqline(resid(model1))

```

# Survey
In studying the drivers of urban housing prices, survey data plays a critical role in understanding household-level factors such as income, housing preferences, and rental demand. To ensure representativeness and minimize bias in the dataset, careful attention must be paid to the sampling methodology employed. 

Urban households in metropolitan areas with a population of over 500,000. The survey aims to sample a diverse range of households across different income levels, family sizes, and housing types.

Sample Size
1,000 households, stratified by:

Income level (low, middle, high)
Housing type (owned, rented, subsidized)
Geographic distribution (central, suburban, peripheral areas)
Survey Method
Online and phone-based surveys, supplemented by in-person interviews for underserved populations.

Survey Questions
Demographic Information

What is the size of your household? (1, 2, 3–4, 5+)

What is your household’s annual income? (Less than $30,000, $30,000–$60,000, $60,000–$100,000, $100,000+)

What is your highest education level? (High school, Associate's, Bachelor's, Graduate)

Housing Information

Do you own or rent your current residence? (Own, Rent, Other)

What is the monthly cost of your housing (including mortgage/rent, utilities)?

Less than $500

$500–$1,000

$1,000–$2,000

$2,000+

What type of housing do you live in? (Apartment, Single-family home, Townhouse, Other)

Housing Market Perceptions

Over the past year, do you feel housing prices in your area have:

Increased significantly

Increased slightly

Remained stable

Decreased

How would you rate the affordability of housing in your area? (1 = Very unaffordable, 5 = Very affordable)
Demand Dynamics

Are you planning to move or upgrade your housing within the next year? (Yes, No, Unsure)

What factors are influencing your decision to move/upgrade? (Check all that apply)

Income changes

Family size changes

Better location

Housing affordability

Job relocation

Additional Context

How far is your residence from your workplace/school? (Less than 5 miles, 5–10 miles, 10–20 miles, 20+ miles)

How much time do you spend commuting daily?

Less than 30 minutes

30–60 minutes

60+ minutes

Ethical Considerations

Ensure participant confidentiality and anonymity.

Provide participants with informed consent forms outlining the purpose of the survey and data usage.

Allow participants to withdraw from the survey at any time.

Data Analysis

The collected data will be used to:

Analyze the relationship between household demographics and housing costs.

Identify factors that influence perceptions of housing affordability.

Model housing price variations across urban areas.

\newpage


# References


